Prey movement shapes the acquisition of predator expertise in a virtual bi-trophic system

### 2024-08-23 manuscript/manuscript.Rmd
update intro. change from development to acquisition.

The acquisition of expertise is crucial for predators to be successful hunters. To achieve this, predators must hone their skills and gain knowledge through repeated and extensive practice. On the other hand, prey may hinder the development of predator expertise by employing antipredator tactics to evade detection and pursuit. However, empirical evidence on how predators acquire expertise through repeated encounters with their prey remains limited, largely due to the challenges of monitoring direct interactions in the wild. Here, we use a virtual predator-prey system (the game *Dead by Daylight*) to investigate how experience shapes individual and population hunting success in human predators across repeated interactions with their prey. We show that predators optimized prey consumption as they gained experience, indicating that they acquired expertise through extensive practice. At the population-level, we found that faster prey impaired the acquisition of expertise by reducing hunting success. Prey speed was also an important mediator of this relationship at the individual level, driving differences among predators in the development of expertise. Our study outlines how prey antipredator behaviour can mediate the acquisition of expertise in predator populations.

@@

Predation is a fundamental biological process involved in ecosystem functioning. Recent advances in our understanding of the mechanisms (e.g. foraging mode, reciprocal plasticity, size-dependent foraging) driving predator foraging success at the individual and population level have allowed ecologists to predict their consequences on prey behaviour and population dynamics [@Peckarsky.etal2008;@gravel.etal2013;@wirsing.etal2021]. Prey typically respond to predation by managing tradeoffs between vigilance and foraging depending on the type of predation regime [@preisser2007]. Such prey antipredator resonses are hypothesized to drive differences among predators in their hunting success by disrupting the acquisition of hunting expertise [@wooster.etal2023], which is deemed essential for many predators to reach adulthood and survive [@woo.etal2008;@wooster.etal2023].

Expertise can be defined as the characteristics, skills, and knowledge allowing individuals with extensive practice to outperform novices on complex tasks [@dukas2017]. In the context of hunting, predators need to acquire expertise by practicing and learning the proper skills to successfully locate, select, and capture their prey. Empirical studies on human and non-human hunters show that individuals optimize foraging efficiency (e.g. search and handling times, return rates) by associative learning, by developing search images, or by exploiting cues from their prey and their environment [@edwards.jackson1994;@morse2000;@macdonald2007;@reid.etal2010;@wilson-rankin2015]. Through these processes, expert predators should have greater knowledge, better energy management, and accute motor skills to increase their chances of locating and capturing prey [@dukas2019]. Therefore, differences among predators in their capacity to acquire expertise throughout their lifetime may underlie differences in hunting success, with potential consequences for predator-prey dynamics [@wooster.etal2023].

Prey are known to use evasive tactics such as rapid escapes and movements to avoid being captured [@walker.etal2005;@kelley.magurran2011;@herbert-read.etal2017], compelling predators to develop swift motor skills to succesfully capture such prey. However, predators may be limited in their capacity to develop the necessary attributes (e.g. physical, physiological, neurological) for fast-paced hunting, which may impair the development of their hunting expertise. There is only scarce evidence showing links between prey antipredator strategies and the development of expertise in human and nonhuman predators. Experimental studies have shown that certain prey camouflage strategies can impair expertise acquisition in humans and birds [@martin.etal2012;@troscianko.etal2013]. For instance, @troscianko.etal2018 found that disruptive coloration interfered with the formation of search images in human subjects, hindering improvement in detection times after repeated attempts. Given these examples, antipredator behaviour may also hold the potential to hinder the development of predator expertise.

A recurring challenge impeding research on predator-prey behavioural interactions is the need to collect data simultaneously on both the predator and the prey. Here we mitigate these challenges by using *Dead by Daylight* as our study system, a videogame where four prey players need to forage for resources while avoiding predation by a fifth player. Similar to agent-based simulations, *Dead by Daylight* provides controlled virtual environments to test ecological hypotheses [see @lymbery.etal2023 for an example with *Age of Empires II*], but with the advantage of having real players that interact in the virtual space. In this game, the predator population is composed of individuals that either ambush or hunt at high speeds (i.e. mean movement speed along a slow-fast continuum), and their success is driven by the movement of the prey [@fraserfranco.etal2022]. The prey can increase their chances of survival by cooperating and moving fast to escape the predator [@cere.etal2021;@fraserfranco.etal2022;@santostefano.etal2024]. The game also ellicits natural reactions in players such as freezing when predation is imminent (personal observations), which corroborates with another virtual ecological study showing that predation drives individual variation in risk perception [@beauchamp2020]. These observations outline how ecological phenomena can emerge from human interactions in virtual systems with fixed rules [@brosnan.etal2017;@kasumovic.etal2017]. Videogames also generate large volumes of data on thousands of interacting players throughout their lifetime in the game under realistic, controlled, and repeatable ecological scenarios. Hence, *Dead by Daylight* allows us to tackle fundamental questions about the role of antipredator behaviour and experience on predator-prey interactions.

In this study, we assess how repeated encounters with prey shapes predator hunting success using data from players in *Dead by Daylight*. First, we investigate how the predator population develops its hunting expertise. We hypothesize that the predator population's success will increase with experience up to a certain level where it will stabilize [@dukas2019]. However, we expect this pattern to change depending on the movement of the prey encountered. We hypothesize that prey will influence the development of expertise, and predict that faster prey will reduce the gain in expertise. Therefore, we then investigate how prey movement influences the development of expertise at the individual level. If prey speed does not influence hunting success, we predict that the gain in expertise will be similar among individuals. Alternatively, if prey speed influences hunting success, then the acquisition of expertise will vary among individuals.

@@

We tested how predators developed their expertise by computing five Bayesian generalized additive mixed models (GAMM) with thin-plate regression splines, all of which estimated the relationship between hunting success (i.e. number of prey consumed) and the predators' cumulative experience (i.e. number of matches played before the current match). We parametrized the models following the method of @pedersen.etal2019. The first model was the simplest, with a common global smoothing function and random intercepts for the predator ID. In this model, we assume that predators have the same development of expertise, with the model estimating a trend for the average individual (i.e. global smoother). For the second model, we included varying individual smoothers for the predator ID. Here, we assume that individual predators share a similar relationship between success and experience, but that this relationship can vary among them (e.g. predator 1 has a steeper curve than predator 2). This enabled us to test whether predators differed in the development of their expertise. In the third model, we kept the individual smoothers for the predators, but removed the global smoother. This model assumes that predators do not share a common relationship between success and experience. The fourth and fifth models were reproductions of the second and third models respectively, where we included the prey speed to assess its effect on the relationship between success and experience. We included the standardized match duration and prey rank as covariates in all five models.

@@

## Development of expertise at the population level

@@

## Development of expertise at the individual level

Prey speed did not influence among individual differences in average hunting success as the posterior distributions of the standard deviations of individual intercepts were almost completely overlapping (Figure 2, median = `r toString(round(sum1[2,]$Estimate, digits = 2))` vs median = `r toString(round(sum2[2,]$Estimate, digits = 2))`). However, individuals differed in the development of their hunting expertise (Figure 1B-D). We found strong evidence that the speed of the prey mediated among individual differences in the linear relationship between success and experience, as there were substantial differences in the standard deviations of the individual slopes between the two models (Figure 2, median = `r toString(round(sum1[3,]$Estimate, digits = 2))` vs median = `r toString(round(sum2[3,]$Estimate, digits = 2))`). Differences among individuals in the direction of the linear relationship between success and experience were `r toString(round(sum1[3,]$Estimate / sum2[3,]$Estimate, digits = 2))` times lower when we removed the effect of prey speed (i.e. accounting for it in the model).

```{r fig1, out.width = "100%", fig.cap = "**Figure 1**. Median posterior predictions of the development of predator hunting expertise. The predators' hunting success (i.e. the probability of consuming the four prey) is on the y axis, and the predators' cumulative experience (i.e. the number of matches played prior to each observation) is on the x axis. Panels A and C show the development of expertise for the average individual with the vertical dashed lines on the left representing the lowest predicted values. For panel A, the right-side vertical dashed line shows the highest predicted success. For panel C, the right-side dashed line represents the point on the curve where success was optimized, which we calculated using the finite differences method to obtain the first derivative of the predicted values. Panels B and D show among individual differences in the development of expertise, with each curve representing an individual predator. (A-B) GAMM where we control for the prey rank (C-D) GAMM where we control for the prey rank and the speed of the prey group."}

@@

Predators usually optimize prey consumption through time by acquiring hunting expertise [@dukas2019;@wooster.etal2023]. Yet, few studies have empirically tested how prey antipredator strategies mediate the acquisition of hunting expertise due to the challenges of investigating direct predator-prey interactions in the wild. By capitalizing on a virtual predator-prey system where interactions were directly monitored, we found that a predator-player population in *Dead by Daylight* increased their hunting success with experience. Yet, there were important differences among individuals in the development of said expertise, which we found to be in part due to the speed of the prey encountered. Hence, our study provides empirical evidence that prey antipredator behaviour can impair the development of hunting expertise.

@@

Prey speed also mediated differences among predator players in the development of their expertise, suggesting that individual predator players differed in their capacity to adjust to difficult prey. Animals are expected to have limited attention that constrain diet choice and image search formation, and hunting faster prey requires specialized cognitive abilities and coordination that are energetically costly [@dukas.etal2001;@kelley.magurran2011]. Thus, predators that couldn't develop counter-strategies to detect or chase faster prey were likely at a disadvantage. Parallel observations have been outlined in studies of prey camouflage strategies. For example, @troscianko.etal2018 showed in a computer experiment involving humans that disruptive camouflage was efficient at preventing the acquisition of expertise during search image formation. Human subjects exposed to a restricted set of strategies were also less efficient compared to those exposed to a variety of strategies. Therefore, our observations suggest that prey antipredator behaviour can also impair predator expertise acquisition, with potential implications for predator-prey dynamics.

@@

We found support of our hypothesis that prey antipredator behaviour was driving individual differences in expertise acquisition in a human predator population in the game *Dead by Daylight*. As we did here with predators, future analyses should investigate how antipredator tactics are developped with experience, as it may reveal important insights on the eco-evolutionary dynamics of predator prey interactions. Using a video game system, our study demonstrates that prey antipredator behavior can impair the acquisition of hunting expertise. This adds to a growing body of research showing how virtual systems can be used to test hypotheses on ecological interactions [@beauchamp2020;@cere.etal2021;@fraserfranco.etal2022;@lymbery.etal2023;@santostefano.etal2024]. We therefore hope that our study will inspire more collaborations between scientists and the videogame industry to tackle fundamental questions in ecology.

>>>>

The acquisition of expertise is crucial for predators to be successful hunters. To achieve this, predators must hone their skills and gain knowledge through repeated and extensive practice. On the other hand, prey may hinder the acquisition of predator expertise by employing antipredator tactics to evade detection and pursuit. However, empirical evidence on how predators acquire expertise through repeated encounters with their prey remains limited, largely due to the challenges of monitoring direct interactions in the wild. Here, we use a virtual predator-prey system (the game *Dead by Daylight*) to investigate how experience shapes individual and population hunting success in human predators across repeated interactions with their prey. We show that predators optimized prey consumption as they gained experience, indicating that they acquired expertise through extensive practice. At the population-level, we found that faster prey impaired the acquisition of expertise by reducing hunting success. Prey speed was also an important mediator of this relationship at the individual level, driving differences among predators in the acquisition of expertise. Our study outlines how prey antipredator behaviour can mediate the acquisition of expertise in predator populations.

@@

Antipredation responses by prey are hypothesized to drive differences among predators in their hunting success by disrupting the acquisition of hunting expertise [@wooster.etal2023]. Expertise can be defined as the characteristics, skills, and knowledge that provide individuals with the ability to outperform novices on complex tasks [@dukas2017]. In the context of hunting, predators need to acquire expertise by practising and learning the proper skills to successfully locate, select, and capture their prey, a process deemed essential for many predators to reach adulthood and survive [@woo.etal2008;@wooster.etal2023]. Empirical studies on human and non-human hunters show that individuals optimize foraging efficiency (e.g. search and handling times, return rates) by associative learning, by developing search images, or by exploiting cues from their prey and their environment [@edwards.jackson1994;@morse2000;@macdonald2007;@reid.etal2010;@wilson-rankin2015]. Through these processes, expert predators should have greater knowledge, better energy management, and acute motor skills to increase their chances of locating and capturing prey [@dukas2019]. Therefore, differences among predators in their capacity to acquire expertise throughout their lifetime may underlie differences in hunting success, with potential consequences for predator-prey dynamics [@wooster.etal2023].

Prey are known to use evasive tactics such as rapid escapes to avoid being captured [@walker.etal2005;@kelley.magurran2011;@herbert-read.etal2017], driving predators to develop the motor skills required to capture such prey succesfully. However, predators may be limited in their capacity to develop the necessary attributes (e.g. physical, physiological, neurological) for fast-paced hunting, which may impair the acquisition of hunting expertise. There is limited empirical evidence showing links between prey antipredator strategies and the acquisition of expertise in human and nonhuman predators. Experimental studies show that certain camouflage strategies of prey can impair expertise acquisition in humans and birds [@martin.etal2012;@troscianko.etal2013]. For instance, @troscianko.etal2018 found that disruptive colouration interfered with the formation of search images in human subjects, hindering improvement in detection times after repeated attempts. Given these examples, antipredator behaviour may also hold the potential to hinder the acquisition of predator expertise. This is an important topic of study since predator cognition can maintain behavioural variation, shape prey phenotypic composition, and destabilize predator-prey systems [@kondoh2010;skelhornandcandy2016;@kikuchi.etal2023].

A recurring challenge impeding research on predator-prey behavioural interactions is the need to collect data simultaneously on both the predator and the prey. Here we mitigate these challenges by using *Dead by Daylight* as our study system, a videogame where four prey players must forage for resources while avoiding predation by a fifth player. Similar to agent-based simulations, *Dead by Daylight* provides controlled virtual environments to test ecological hypotheses [see @lymbery.etal2023 for an example with *Age of Empires II*], but with the advantage of having real players that interact in the virtual space. In this game, the predator population comprises individuals that either ambush or hunt at high speeds (i.e. mean movement speed along a slow-fast continuum), and their success is driven by the movement of the prey [@fraserfranco.etal2022]. The prey can increase their chances of survival by cooperating and moving fast to escape the predator [@cere.etal2021;@fraserfranco.etal2022;@santostefano.etal2024]. The game also elicits natural reactions in players such as freezing when predation is imminent (personal observations), corroborating another virtual ecological study showing that predation drives individual variation in risk perception [@beauchamp2020]. These observations outline how ecological phenomena can emerge from human interactions in virtual systems with fixed rules [@brosnan.etal2017;@kasumovic.etal2017]. Videogames also generate large volumes of data on thousands of interacting players throughout their lifetime in the game under realistic, controlled, and repeatable ecological scenarios. Hence, *Dead by Daylight* allows us to tackle fundamental questions about the role of antipredator behaviour and experience in predator-prey interactions.

In this study, we assess how repeated encounters with prey shapes predator hunting success using data from players in *Dead by Daylight*. We quantify expertise acquisition as the relationship between hunting success (i.e. the probability of capturing all prey) and repeated experience (i.e. cumulated atempts). First, test the hypothesis that predator success will increase with experience up to some point at which it will stabilize [@dukas2019]. However, we expect this pattern to change depending on the movement of the prey encountered. We hypothesize that prey will influence expertise acquisition, and predict that faster prey will reduce the gain in expertise. Therefore, we also investigate how prey movement influences expertise acquisition at the individual level. If prey speed does not influence hunting success, we predict that the gain in expertise will be similar among individuals. Alternatively, if prey speed influences hunting success, then the acquisition of expertise will vary among individuals.

@@

We tested how predators developed their expertise by computing five Bayesian generalized additive mixed models (GAMM) with thin-plate regression splines, all of which estimated the relationship between hunting success (i.e. number of prey consumed) and the predators' cumulative experience (i.e. number of matches played before the current match). We parametrized the models following the method of @pedersen.etal2019. The first model was the simplest, with a common global smoothing function and random intercepts for the predator ID. In this model, we assume that predators have the same acquisition of expertise, with the model estimating a trend for the average individual (i.e. global smoother). For the second model, we included varying individual smoothers for the predator ID. Here, we assume that individual predators share a similar relationship between success and experience, but that this relationship can vary among them (e.g. predator 1 has a steeper curve than predator 2). This enabled us to test whether predators differed in the development of their expertise. In the third model, we kept the individual smoothers for the predators, but removed the global smoother. This model assumes that predators do not share a common relationship between success and experience. The fourth and fifth models were reproductions of the second and third models respectively, where we included the prey speed to assess its effect on the relationship between success and experience. We included the standardized match duration and prey rank as covariates in all five models.

@@

## Acquisition of expertise at the population level

@@

## Acquisition of expertise at the individual level

Prey speed did not influence among individual differences in average hunting success as the posterior distributions of the standard deviations of individual intercepts were almost completely overlapping (Figure 2, median = `r toString(round(sum1[2,]$Estimate, digits = 2))` vs median = `r toString(round(sum2[2,]$Estimate, digits = 2))`). However, individuals differed in the acquisition of their hunting expertise (Figure 1B-D). We found strong evidence that the speed of the prey mediated among individual differences in the linear relationship between success and experience, as there were substantial differences in the standard deviations of the individual slopes between the two models (Figure 2, median = `r toString(round(sum1[3,]$Estimate, digits = 2))` vs median = `r toString(round(sum2[3,]$Estimate, digits = 2))`). Differences among individuals in the direction of the linear relationship between success and experience were `r toString(round(sum1[3,]$Estimate / sum2[3,]$Estimate, digits = 2))` times lower when we removed the effect of prey speed (i.e. accounting for it in the model).

```{r fig1, out.width = "100%", fig.cap = "**Figure 1**. Median posterior predictions of the acquisition of predator hunting expertise. The predators' hunting success (i.e. the probability of consuming the four prey) is on the y axis, and the predators' cumulative experience (i.e. the number of matches played prior to each observation) is on the x axis. Panels A and C show the acquisition of expertise for the average individual with the vertical dashed lines on the left representing the lowest predicted values. For panel A, the right-side vertical dashed line shows the highest predicted success. For panel C, the right-side dashed line represents the point on the curve where success was optimized, which we calculated using the finite differences method to obtain the first derivative of the predicted values. Panels B and D show among individual differences in the acquisition of expertise, with each curve representing an individual predator. (A-B) GAMM where we control for the prey rank (C-D) GAMM where we control for the prey rank and the speed of the prey group."}

@@

Predators usually optimize prey consumption through time by acquiring hunting expertise [@dukas2019;@wooster.etal2023]. Yet, few studies have empirically tested how prey antipredator strategies mediate the acquisition of hunting expertise due to the challenges of investigating direct predator-prey interactions in the wild. By capitalizing on a virtual predator-prey system where interactions were directly monitored, we found that a predator-player population in *Dead by Daylight* increased their hunting success with experience. Yet, there were important differences among individuals in the acquisition of said expertise, which we found to be in part due to the speed of the prey encountered. Hence, our study provides empirical evidence that prey antipredator behaviour can impair the acquisition of hunting expertise.

@@

Prey speed also mediated differences among predator players in the acquisition of their expertise, suggesting that individual predator players differed in their capacity to adjust to difficult prey. Animals are expected to have limited attention that constrain diet choice and image search formation, and hunting faster prey requires specialized cognitive abilities and coordination that are energetically costly [@dukas.etal2001;@kelley.magurran2011]. Thus, predators that couldn't develop counter-strategies to detect or chase faster prey were likely at a disadvantage. Parallel observations have been outlined in studies of prey camouflage strategies. For example, @troscianko.etal2018 showed in a computer experiment involving humans that disruptive camouflage was efficient at preventing the acquisition of expertise during search image formation. Human subjects exposed to a restricted set of strategies were also less efficient compared to those exposed to a variety of strategies. Therefore, our observations suggest that prey antipredator behaviour can also impair predator expertise acquisition, with potential implications for predator-prey dynamics.

@@

We found support of our hypothesis that prey antipredator behaviour was driving individual differences in expertise acquisition in a human predator population in the game *Dead by Daylight*. As we did here with predators, future analyses should investigate how antipredator tactics are developped with experience, as it may reveal important insights on the eco-evolutionary dynamics of predator prey interactions. Using a video game system, our study demonstrates that prey antipredator behaviour can impair the acquisition of hunting expertise. This adds to a growing body of research showing how virtual systems can be used to test hypotheses on ecological interactions [@beauchamp2020;@cere.etal2021;@fraserfranco.etal2022;@lymbery.etal2023;@santostefano.etal2024]. We therefore hope that our study will inspire more collaborations between scientists and the videogame industry to tackle fundamental questions in ecology.
